---
title: "AgingJobUpdate"
author: "Obydah Jazaeri, Shahlan Finchan, Sarah Rosen"
date: "2026-02-26"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(writexl)
library(tidyverse)
library(janitor)    
library(readxl)
library(tidyr)
library(stringr) 
library(ggplot2)
library(plotly)
library(dplyr)
library(scales)
knitr::opts_chunk$set(echo = TRUE)
data2011 <- read_excel("11b_2011.xlsx")
data2012 <- read_excel("11b_2012.xlsx")
data2013 <- read_excel("11b_2013.xlsx")
data2014 <- read_excel("11b_2014.xlsx")
data2015 <- read_excel("11b_2015.xlsx")
data2016 <- read_excel("11b_2016.xlsx")
data2017 <- read_excel("11b_2017.xlsx")
data2018 <- read_excel("11b_2018.xlsx")
data2019 <- read_excel("11b_2019.xlsx")
data2020 <- read_excel("11b_2020.xlsx")
data2021 <- read_excel("11b_2021.xlsx")
data2022 <- read_excel("11b_2022.xlsx")
data2023 <- read_excel("11b_2023.xlsx")
data2024 <- read_excel("11b_2024.xlsx")
table8_data <- read_excel("Table_8_Data.xlsx")
```

```{r cleaning, include=FALSE}
clean_cps_year <- function(df, year) {
  
  df %>%
    slice(-c(1:3)) %>%   
    rename(
      occupation = 1,
      median_age = 2,
      `16_to_19_years` = 3,
      `20_to_24_years` = 4,
      `25_to_34_years` = 5,
      `35_to_44_years` = 6,
      `45_to_54_years` = 7,
      `55_to_64_years` = 8,
      `65_years_and_over` = 9
    ) %>%
    filter(occupation != "Total employed") %>%
    
    # DROP median age before pivoting
    select(-median_age) %>%
    
    pivot_longer(
      cols = -occupation,
      names_to = "age_group",
      values_to = "employment_thousands"
    ) %>%
    mutate(
      year = year,
      employment_thousands = readr::parse_number(employment_thousands)
    )
}

```

```{r create-cps-clean, include=FALSE}
data2011_clean <- clean_cps_year(data2011, 2011)
data2012_clean <- clean_cps_year(data2012, 2012)
data2013_clean <- clean_cps_year(data2013, 2013)
data2014_clean <- clean_cps_year(data2014, 2014)
data2015_clean <- clean_cps_year(data2015, 2015)
data2016_clean <- clean_cps_year(data2016, 2016)
data2017_clean <- clean_cps_year(data2017, 2017)
data2018_clean <- clean_cps_year(data2018, 2018)
data2019_clean <- clean_cps_year(data2019, 2019)
data2020_clean <- clean_cps_year(data2020, 2020)
data2021_clean <- clean_cps_year(data2021, 2021)
data2022_clean <- clean_cps_year(data2022, 2022)
data2023_clean <- clean_cps_year(data2023, 2023)
data2024_clean <- clean_cps_year(data2024, 2024
)

                 
clean11b_data <- bind_rows(
  data2011_clean,
  data2012_clean,
  data2013_clean,
  data2014_clean,
  data2015_clean,
  data2016_clean,
  data2017_clean,
  data2018_clean,
  data2019_clean,
  data2020_clean,
  data2021_clean,
  data2022_clean,
  data2023_clean,
  data2024_clean
)
```


```{r retirement-prep, include=FALSE}
retire_metrics <- clean11b_data %>%
  group_by(occupation, year) %>%
  summarise(
    c65 = sum(employment_thousands[age_group == "65_years_and_over"], na.rm = TRUE),
    c55 = sum(employment_thousands[age_group %in% c("55_to_64_years", "65_years_and_over")], na.rm = TRUE),
    total = sum(employment_thousands, na.rm = TRUE),
    pct65 = 100 * c65 / total,
    pct55 = 100 * c55 / total,
    .groups = "drop"
  ) %>%
  filter(!is.na(total), total > 0)

occ_choices_ret <- sort(unique(retire_metrics$occupation))
year_min <- min(retire_metrics$year, na.rm = TRUE)
year_max <- max(retire_metrics$year, na.rm = TRUE)
```


```{r Shahlans Cleaning Code for 8, include=FALSE}
t8 <- read_excel(
  "Table_8_Data.xlsx",
  col_types = c("text","text","text","text","text","text","text","text")
)
glimpse(t8)

# Convert all columns from text to numbers
t8 <- t8 %>%
  mutate(across(c(FT_Total, PT_Total, Unemp_FT, Unemp_PT), readr::parse_number))

# Pivot to Tidy Format
t8_tidy <- t8 %>%
  pivot_longer(
    cols = c(FT_Total, PT_Total, Unemp_FT, Unemp_PT),
    names_to = c("Measure","Time"),
    names_sep = "_",
    values_to = "Count"
  ) %>%
  # Combine Measure and Time info into meaningful columns
  mutate(
    FT = if_else(Measure == "FT" & Time == "Total", Count, NA_real_),
    PT = if_else(Measure == "PT" & Time == "Total", Count, NA_real_),
    Unemp = if_else(Measure == "Unemp", Count, NA_real_)
  ) %>%
  group_by(Year, Sex, Race, Age) %>%
  summarise(
    FT = sum(FT, na.rm = TRUE),
    PT = sum(PT, na.rm = TRUE),
    Unemp = sum(Unemp, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(Total = FT + PT)


glimpse(t8_tidy)



t8_tidy <- t8_tidy %>%
  filter(!is.na(Sex) & !is.na(Race) & !is.na(Age))


t8_tidy <- t8_tidy %>%
  mutate(
    Total = as.numeric(Total),
    Unemp = as.numeric(Unemp)
  )

t8_tidy <- t8_tidy %>%
  mutate(Year = as.numeric(Year))


t8_tidy <- t8_tidy %>%
  mutate(Age = factor(Age, 
                      levels = c("16 to 19","20 to 24","20+","25 to 54","25+","55+"),
                      ordered = TRUE))


write_xlsx(t8_tidy, "Table_8_Data_Tidy.xlsx")

summary(t8_tidy)

clean_table8 <- function(df) {
  
  df %>%
    
    pivot_longer(
      cols = c(FT_Total, PT_Total, Unemp_FT, Unemp_PT),
      names_to = c("measure","time"),
      names_sep = "_",
      values_to = "count"
    ) %>%
    
   
    mutate(
      FT = if_else(measure == "FT" & time == "Total", count, NA_real_),
      PT = if_else(measure == "PT" & time == "Total", count, NA_real_),
      Unemp = if_else(measure == "Unemp", count, NA_real_)
    ) %>%
    
    
    group_by(Year, Sex, Race, Age) %>%
    summarise(
      FT = sum(FT, na.rm = TRUE),
      PT = sum(PT, na.rm = TRUE),
      Unemp = sum(Unemp, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    
    
  mutate(
  Total = FT + PT,
  Year = as.integer(Year),
  Age = factor(
    Age,
    levels = c("16 to 19","20 to 24","20+","25 to 54","25+","55+"),
    ordered = TRUE
  )
) %>%

filter(
  !is.na(Sex),
  !is.na(Race),
  !is.na(Age),
  !Age %in% c("20+", "25+")
)
}


clean8_data <- clean_table8(table8_data)

```




```{r Shahlans Graph Section, echo=FALSE}
library(shiny)
library(ggplot2)
library(dplyr)
library(plotly)
library(ggthemes)
library(scales)

# Define colors
myColors <- c(
  "#264653",  
  "#2A9D8F",  
  "#E9C46A", 
  "#E76F51",  
  "#6D597A",
  "#f542b0"
)

# Prepare summarized data for plotting
t8_unemp_age <- clean8_data %>%
  filter(!Age %in% c("20+", "25+")) %>%
  mutate(
    Labor_Force = Total + Unemp,
    Unemployment_Rate = Unemp / Labor_Force
  ) %>%
  group_by(Year, Age) %>%
  summarise(
    Unemployment_Rate = mean(Unemployment_Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ungroup() %>%
  droplevels()

# Shiny UI
ui <- fluidPage(
  titlePanel("Unemployment Rate by Age Group"),
  
  sidebarLayout(
    sidebarPanel(
      checkboxGroupInput(
        "age_select",
        "Select Age Groups:",
        choices = levels(droplevels(t8_unemp_age$Age)),
        selected = levels(droplevels(t8_unemp_age$Age))
      ),
      selectInput(
        inputId = "top_n",
        label = "Number of Top Occupations:",
        choices = c(3, 5, 10, 15),
        selected = 5
    ),
      sliderInput(
        "year_range",
        "Select Year Range:",
        min = min(t8_unemp_age$Year),
        max = max(t8_unemp_age$Year),
        value = c(min(t8_unemp_age$Year), max(t8_unemp_age$Year)),
        step = 1,
        sep = ""
      ),
      checkboxInput("show_points", "Show Points", value = TRUE)
    ),
    
    mainPanel(
      plotlyOutput("unempPlot")
    )
  )
)

# Shiny Server
server <- function(input, output) {
  
  output$piePlot <- renderPlot({
    
    clean11b_data %>%
      dplyr::filter(year == input$year) %>%
      dplyr::group_by(occupation) %>%
      dplyr::summarise(
        total_employment = sum(employment_thousands, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      dplyr::arrange(dplyr::desc(total_employment)) %>%
      dplyr::slice_head(n = as.numeric(input$top_n)) %>%
      ggplot(aes(x = 1, y = total_employment, fill = occupation)) +
      geom_col(width = 1) +
      coord_polar(theta = "y") +
      labs(
        title = paste(
          "Top", input$top_n,
          "Occupations by Employment,", input$year
        ),
        fill = "Occupation",
  filtered_data <- reactive({
    t8_unemp_age %>%
      filter(
        Age %in% input$age_select,
        Year >= input$year_range[1],
        Year <= input$year_range[2]
      )
  }))
  })
}
  
  output$unempPlot <- renderPlotly({
    
    p <- ggplot(
      filtered_data(),
      aes(
        x = Year,
        y = Unemployment_Rate,
        color = Age,
        group = Age,
        text = paste(
          "Year:", Year,
          "<br>Age:", Age,
          "<br>Unemployment Rate:", percent(Unemployment_Rate, accuracy = 1)
        )
      )
    ) +
      geom_line(linewidth = 1.2, alpha = 0.8) +
      scale_color_manual(values = myColors) +
      scale_y_continuous(labels = percent_format()) +
      labs(
        title = "Unemployment Rate by Age Group (2011–2024)",
        subtitle = "Comparing unemployment trends across age groups",
        x = NULL,
        y = NULL,
        color = "Age Group"
      ) +
      theme_fivethirtyeight() +
      theme(text = element_text(family = "Times New Roman"))
    
    if (input$show_points) {
      p <- p + geom_point(size = 2)
    }
    
    ggplotly(p, tooltip = "text") %>%
      layout(
        xaxis = list(fixedrange = TRUE),
        yaxis = list(fixedrange = TRUE),
        hoverlabel = list(
          font = list(family = "Times New Roman", size = 12, color = "black"),
          bgcolor = "white",
          bordercolor = "transparent"
        )
      ) %>%
      config(displayModeBar = FALSE)
  })

shinyApp(ui = ui, server = server )
```

# Run the Shiny App
shinyApp(ui, server)

<<<<<<< HEAD
```



=======
```{r Obydahs Graph section, echo=FALSE}

=======
```{r Obydahs Graph section, echo=FALSE}
>>>>>>> 70dfcf191f66000ce52d600a512cda30af5d5484
ui <- fluidPage(
  
  titlePanel("Job Distribution by Occupation"),
  
  sidebarLayout(
    
    sidebarPanel(
      
      selectInput(
        inputId = "year",
        label = "Select Year:",
        choices = sort(unique(clean11b_data$year)),
        selected = max(clean11b_data$year)
      ),
      
      selectInput(
        inputId = "top_n",
        label = "Number of Top Occupations:",
        choices = c(3, 5, 10, 15),
        selected = 5
      )
    ),
    
    mainPanel(
      plotOutput("piePlot")
    )
  )
)

server <- function(input, output) {
  
  output$piePlot <- renderPlot({
    
    clean11b_data %>%
      dplyr::filter(year == input$year) %>%
      dplyr::group_by(occupation) %>%
      dplyr::summarise(
        total_employment = sum(employment_thousands, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      dplyr::arrange(dplyr::desc(total_employment)) %>%
      dplyr::slice_head(n = as.numeric(input$top_n)) %>%
      ggplot(aes(x = 1, y = total_employment, fill = occupation)) +
      geom_col(width = 1) +
      coord_polar(theta = "y") +
      labs(
        title = paste(
          "Top", input$top_n,
          "Occupations by Employment,", input$year
        ),
        fill = "Occupation"
      ) +
      theme_void()
  })
}

shinyApp (ui = ui, server = server )
```


```{r Dareens Graph Section, echo = FALSE}

  # Shiny UI

  ui <- fluidPage(
    titlePanel("Entry Age Level Occupations by Year"),
  
    sidebarLayout(
      
       sidebarPanel(
      
          selectInput("Select_FirstYear", 
                  "Select First Year", 
                  choices = unique(clean11b_data$year),
                  selected = "2011"
                  ),
          
          selectInput("Select_SecondYear", 
                 "Select Second Year", 
                  choices = unique(clean11b_data$year),
                  selected = "2024"
                ), 
         
          selectInput("Highlight_Occupation",
                  "Select/Type Occupation to Highlight", 
                  choices = c("None", unique(clean11b_data$occupation)), 
                  selected = "None"
                  ),
          
        radioButtons("Select_TopOccupations", 
                  "Select Top N Occupations",
                  choices = c(3, 5, 10, 20, 30), 
                  selected = 3
                  )
    
  ),
    
    mainPanel(
      
     tabsetPanel(
       tabPanel("Graph", plotlyOutput("Entry_Age_Occupation_Map1")),
       tabPanel("Table", tableOutput("Top_Occupations_Table1"))
     ),
     
     tabsetPanel(
       tabPanel("Graph", plotlyOutput("Entry_Age_Occupation_Map2")),
       tabPanel("Table", tableOutput("Top_Occupations_Table2") )
     )
    )
  )
)

  # Shiny Server
 server <- function(input, output) {
  
  # Function to prepare data 
  get_top_occupation <- function(selected_year){ 
    
    data <- clean11b_data %>% 
    filter(age_group == "20_to_24_years", 
           year == selected_year) %>%
    slice_max(employment_thousands,
              n = as.numeric(input$Select_TopOccupations))
  
  data %>%
    mutate(highlight = ifelse(occupation == input$Highlight_Occupation, 
                              "Selected", 
                              "Other")
           )
  }
  
  #Function to Create Plots 
  make_plot <- function(data, selected_year){
    p <- data %>%
      ggplot(aes(x = reorder(occupation, employment_thousands),
                 y = employment_thousands,
                 fill = if(input$Highlight_Occupation == "None"){
                   occupation
                 } else {
                   highlight
                 }, 
                 text = paste(
                   "Occupation:", occupation, 
                   "<br> Total Employment:", employment_thousands, "(thousands)",
                   "<br> Year", year)
              
                 )
             ) +
      
      geom_col(width = 0.7) +
      labs(title = paste0(" Top ", input$Select_TopOccupations, " Entry Age Occupations ", "(",selected_year,")"),
           x = "Occupation",
           y = "Total Employment") +
      coord_flip() +
      theme_minimal() +
      theme(legend.position = "none") +
      theme(plot.title = element_text(
          face = "bold.italic", 
          size = 16
          )
      )
    
    
    # If User Selected an Occupation Highlight it with red and others as gray 
    if(input$Highlight_Occupation != "None"){
      p <- p + scale_fill_manual(values = c(
        "Selected" = "red", 
        "Other" = "gray"))
    }
    
    ggplotly(p, tooltip = "text")
  }
  
  # Graph 1
  output$Entry_Age_Occupation_Map1 <- renderPlotly({ 
    data <- get_top_occupation(input$Select_FirstYear) 
    
      make_plot(data, input$Select_FirstYear)
    })
  
  # Graph 2
  output$Entry_Age_Occupation_Map2 <- renderPlotly({
    data <- get_top_occupation(input$Select_SecondYear)
    
    make_plot(data, input$Select_SecondYear)
  })
  
  # Table 1
  output$Top_Occupations_Table1 <- renderTable({
    clean11b_data %>%
      filter(age_group == "20_to_24_years",
             year == input$Select_FirstYear) %>%
      slice_max(employment_thousands, n = as.numeric(input$Select_TopOccupations))
  })
  
  # Table 2
  output$Top_Occupations_Table2 <- renderTable({
    clean11b_data %>%
    filter(age_group == "20_to_24_years", 
           year == input$Select_SecondYear) %>%
      slice_max(employment_thousands, n = as.numeric(input$Select_TopOccupations))
  })
 }
 
shinyApp(ui, server)

```

## Retirement Projections

This chart tracks how “near-retirement” employment changes over time for a selected occupation. Users can switch between counts and percentages for ages 55+ and 65+ to compare retirement pressure across jobs. The dashed extension is a simple linear projection to help visualize where recent trends may be heading (an exploratory forecast, not a guarantee).

```{r Sarahs Graphs Section, echo=FALSE}
shiny::fluidRow(
  shiny::column(
    4,
    shiny::selectizeInput(
      "occ_ret",
      "Choose an occupation:",
      choices = occ_choices_ret,
      selected = occ_choices_ret[1],
      options = list(placeholder = "Type to search...", maxOptions = 5000)
    ),
    shiny::radioButtons(
      "metric_ret",
      "Metric:",
      choices = c(
        "Percent age 55+ (retirement pressure)" = "pct55",
        "Percent age 65+ (near/at retirement)"  = "pct65",
        "Count age 55+ (thousands)" = "c55",
        "Count age 65+ (thousands)" = "c65"
      ),
      selected = "pct55"
    ),
    shiny::sliderInput(
      "year_ret",
      "Year range:",
      min = year_min, max = year_max,
      value = c(year_min, year_max),
      step = 1, sep = ""
    ),
    shiny::checkboxInput("show_proj", "Add simple projection (linear trend)", value = TRUE),
    shiny::sliderInput("proj_years", "Projection years", min = 1, max = 10, value = 5)
  ),
  shiny::column(
    8,
    plotly::plotlyOutput("retire_line", height = "450px"),
    shiny::tags$hr(),
    shiny::textOutput("retire_interpretation"),
    shiny::helpText("Tip: Drag to zoom, double-click to reset. Hover for exact values.")
  )
)

metric_data <- reactive({
  req(input$occ_ret)

  retire_metrics %>%
    filter(
      occupation == input$occ_ret,
      year >= input$year_ret[1],
      year <= input$year_ret[2]
    ) %>%
    arrange(year) %>%
    mutate(
      value = dplyr::case_when(
        input$metric_ret == "pct55" ~ pct55,
        input$metric_ret == "pct65" ~ pct65,
        input$metric_ret == "c55"   ~ c55,
        input$metric_ret == "c65"   ~ c65,
        TRUE ~ pct55
      ),
      metric_label = dplyr::case_when(
        input$metric_ret == "pct55" ~ "% age 55+",
        input$metric_ret == "pct65" ~ "% age 65+",
        input$metric_ret == "c55"   ~ "Workers age 55+ (thousands)",
        input$metric_ret == "c65"   ~ "Workers age 65+ (thousands)",
        TRUE ~ "% age 55+"
      )
    )
})
output$retire_line <- plotly::renderPlotly({
  d <- metric_data()
  req(nrow(d) > 0)

  p <- ggplot(d, aes(x = year, y = value)) +
    geom_line(color = "#2C7FB8", size = 1.2) +
    geom_point(color = "#2C7FB8", size = 2) +
    labs(
      title = paste("Retirement Trend:", input$occ_ret),
      x = "Year",
      y = unique(d$metric_label)
    ) +
    theme_minimal(base_size = 13)

  if (isTRUE(input$show_proj) && nrow(d) >= 3 && all(is.finite(d$value))) {
    fit <- lm(value ~ year, data = d)
    last_year <- max(d$year, na.rm = TRUE)
    future_years <- (last_year + 1):(last_year + input$proj_years)

    proj <- tibble(
      year = future_years,
      value = predict(fit, newdata = tibble(year = future_years))
    )

    p <- p +
      geom_line(data = proj, aes(year, value), linetype = "dashed") +
      geom_point(data = proj, aes(year, value))
  }

  plotly::ggplotly(p, tooltip = c("x", "y")) %>% 
  plotly::config(displayModeBar = TRUE) %>%
  plotly::layout(
    hovermode = "closest",
    legend = list(
      orientation = "v",
      bgcolor = "rgba(255,255,255,0.7)",
      x = 1.05,
      xanchor = "left",
      y = 1,
      yanchor = "top"
    ),
    margin = list(t = 80)
  )

  ggplotly(p, tooltip = c("x", "y")) %>%
    layout(
      hovermode = "closest",
      legend = list(orientation = "v")
    )
})
```

```{r Sammys Graph Section}

```


```{r Rishitas Graph Section, echo=FALSE}

ui <- fluidPage(
  titlePanel("Unemployment Rate by Race (2011–2023)"),
  sidebarLayout(
    sidebarPanel(
      checkboxGroupInput(
        "race_select",
        "Select Race Groups:",
        choices = unique(clean8_data$Race),
        selected = unique(clean8_data$Race)
      ),
      sliderInput(
        "year_range_race",
        "Select Year Range:",
        min = min(clean8_data$Year),
        max = max(clean8_data$Year),
        value = c(min(clean8_data$Year), max(clean8_data$Year)),
        step = 1,
        sep = ""
      )
    ),
    mainPanel(
      plotlyOutput("racePlot")
    )
  )
)

server <- function(input, output) {
  race_data <- reactive({
    clean8_data %>%
      filter(
        Race %in% input$race_select,
        Year >= input$year_range_race[1],
        Year <= input$year_range_race[2]
      ) %>%
      mutate(
        Labor_Force = Total + Unemp,
        Unemployment_Rate = Unemp / Labor_Force
      ) %>%
      group_by(Year, Race) %>%
      summarise(
        Unemployment_Rate = mean(Unemployment_Rate, na.rm = TRUE),
        .groups = "drop"
      )
  })
  
  output$racePlot <- renderPlotly({
    p <- ggplot(
      race_data(),
      aes(
        x = Year,
        y = Unemployment_Rate,
        color = Race,
        group = Race,
        text = paste(
          "Year:", Year,
          "<br>Race:", Race,
          "<br>Unemployment Rate:", percent(Unemployment_Rate, accuracy = 0.1)
        )
      )
    ) +
      geom_line(size = 1.3) +
      geom_point(size = 2) +
      scale_y_continuous(labels = percent_format()) +
      labs(
        title = "Unemployment Rate Trends by Race",
        subtitle = "Comparing racial groups across time",
        x = "Year",
        y = "Unemployment Rate",
        color = "Race"
      ) +
      theme_minimal(base_size = 14)
    ggplotly(p, tooltip = "text") %>%
      layout(
        hovermode = "closest"
      ) %>%
      config(displayModeBar = FALSE)
  })
}

shinyApp (ui = ui, server = server)
```

## Retirement

This visualization focuses on workers age 55 and older to look at retirement-related trends. By tracking employment rates over time, we can see how participation in the workforce among older workers has changed from 2011 to 2024.

```{r Zuwiyda Graph Section}

```







```