---
title: "AgingJobUpdate"
author: "Obydah Jazaeri, Shahlan Finchan"
date: "2026-02-26"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(writexl)
library(tidyverse)
library(janitor)    
library(readxl)
library(tidyr)
library(stringr) 
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
data2011 <- read_excel("11b_2011.xlsx")
data2012 <- read_excel("11b_2012.xlsx")
data2013 <- read_excel("11b_2013.xlsx")
data2014 <- read_excel("11b_2014.xlsx")
data2015 <- read_excel("11b_2015.xlsx")
data2016 <- read_excel("11b_2016.xlsx")
data2017 <- read_excel("11b_2017.xlsx")
data2018 <- read_excel("11b_2018.xlsx")
data2019 <- read_excel("11b_2019.xlsx")
data2020 <- read_excel("11b_2020.xlsx")
data2021 <- read_excel("11b_2021.xlsx")
data2022 <- read_excel("11b_2022.xlsx")
data2023 <- read_excel("11b_2023.xlsx")
data2024 <- read_excel("11b_2024.xlsx")
table8_data <- read_excel("Table_8_Data.xlsx")
```

```{r cleaning}
clean_cps_year <- function(df, year) {
  
  df %>%
    slice(-c(1:3)) %>%   
    rename(
      occupation = 1,
      median_age = 2,
      `16_to_19_years` = 3,
      `20_to_24_years` = 4,
      `25_to_34_years` = 5,
      `35_to_44_years` = 6,
      `45_to_54_years` = 7,
      `55_to_64_years` = 8,
      `65_years_and_over` = 9
    ) %>%
    filter(occupation != "Total employed") %>%
    
    # DROP median age before pivoting
    select(-median_age) %>%
    
    pivot_longer(
      cols = -occupation,
      names_to = "age_group",
      values_to = "employment_thousands"
    ) %>%
    mutate(
      year = year,
      employment_thousands = readr::parse_number(employment_thousands)
    )
}

```

```{r create-cps-clean}
data2011_clean <- clean_cps_year(data2011, 2011)
data2012_clean <- clean_cps_year(data2012, 2012)
data2013_clean <- clean_cps_year(data2013, 2013)
data2014_clean <- clean_cps_year(data2014, 2014)
data2015_clean <- clean_cps_year(data2015, 2015)
data2016_clean <- clean_cps_year(data2016, 2016)
data2017_clean <- clean_cps_year(data2017, 2017)
data2018_clean <- clean_cps_year(data2018, 2018)
data2019_clean <- clean_cps_year(data2019, 2019)
data2020_clean <- clean_cps_year(data2020, 2020)
data2021_clean <- clean_cps_year(data2021, 2021)
data2022_clean <- clean_cps_year(data2022, 2022)
data2023_clean <- clean_cps_year(data2023, 2023)
data2024_clean <- clean_cps_year(data2024, 2024
)

                 
clean11b_data <- bind_rows(
  data2011_clean,
  data2012_clean,
  data2013_clean,
  data2014_clean,
  data2015_clean,
  data2016_clean,
  data2017_clean,
  data2018_clean,
  data2019_clean,
  data2020_clean,
  data2021_clean,
  data2022_clean,
  data2023_clean,
  data2024_clean
)
```

<<<<<<< HEAD
```{r Shahlans Cleaning Code for 8}
t8 <- read_excel(
  "Table_8_Data.xlsx",
  col_types = c("text","text","text","text","text","text","text","text")
)
glimpse(t8)

# Convert all columns from text to numbers
t8 <- t8 %>%
  mutate(across(c(FT_Total, PT_Total, Unemp_FT, Unemp_PT), readr::parse_number))

# Pivot to Tidy Format
t8_tidy <- t8 %>%
  pivot_longer(
    cols = c(FT_Total, PT_Total, Unemp_FT, Unemp_PT),
    names_to = c("Measure","Time"),
    names_sep = "_",
    values_to = "Count"
  ) %>%
  # Combine Measure and Time info into meaningful columns
  mutate(
    FT = if_else(Measure == "FT" & Time == "Total", Count, NA_real_),
    PT = if_else(Measure == "PT" & Time == "Total", Count, NA_real_),
    Unemp = if_else(Measure == "Unemp", Count, NA_real_)
  ) %>%
  group_by(Year, Sex, Race, Age) %>%
  summarise(
    FT = sum(FT, na.rm = TRUE),
    PT = sum(PT, na.rm = TRUE),
    Unemp = sum(Unemp, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(Total = FT + PT)


glimpse(t8_tidy)



t8_tidy <- t8_tidy %>%
  filter(!is.na(Sex) & !is.na(Race) & !is.na(Age))


t8_tidy <- t8_tidy %>%
  mutate(
    Total = as.numeric(Total),
    Unemp = as.numeric(Unemp)
  )

t8_tidy <- t8_tidy %>%
  mutate(Year = as.numeric(Year))


t8_tidy <- t8_tidy %>%
  mutate(Age = factor(Age, 
                      levels = c("16 to 19","20 to 24","20+","25 to 54","25+","55+"),
                      ordered = TRUE))


write_xlsx(t8_tidy, "Table_8_Data_Tidy.xlsx")

summary(t8_tidy)

clean_table8 <- function(df) {
  
  df %>%
    
    pivot_longer(
      cols = c(FT_Total, PT_Total, Unemp_FT, Unemp_PT),
      names_to = c("measure","time"),
      names_sep = "_",
      values_to = "count"
    ) %>%
    
   
    mutate(
      FT = if_else(measure == "FT" & time == "Total", count, NA_real_),
      PT = if_else(measure == "PT" & time == "Total", count, NA_real_),
      Unemp = if_else(measure == "Unemp", count, NA_real_)
    ) %>%
    
    
    group_by(Year, Sex, Race, Age) %>%
    summarise(
      FT = sum(FT, na.rm = TRUE),
      PT = sum(PT, na.rm = TRUE),
      Unemp = sum(Unemp, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    
    
  mutate(
  Total = FT + PT,
  Year = as.integer(Year),
  Age = factor(
    Age,
    levels = c("16 to 19","20 to 24","20+","25 to 54","25+","55+"),
    ordered = TRUE
  )
) %>%

filter(
  !is.na(Sex),
  !is.na(Race),
  !is.na(Age),
  !Age %in% c("20+", "25+")
)
}


clean8_data <- clean_table8(table8_data)

```


```{r Obydahs Graph section}

```


```{r Shahlans Graph Section}

library(shiny)
library(ggplot2)
library(dplyr)
library(ggthemes)
library(plotly)
library(scales)


myColors <- c(
  "#264653",  
  "#2A9D8F",  
  "#E9C46A", 
  "#E76F51",  
  "#6D597A"   
)

ui <- fluidPage(
  
  titlePanel("Unemployment Rate by Age Group"),
  
  sidebarLayout(
    
    sidebarPanel(
      
      checkboxGroupInput(
        "age_select",
        "Select Age Groups:",
        choices = unique(t8_unemp_age$Age),
        selected = unique(t8_unemp_age$Age)
      ),
      checkboxInput("show_points", "Show Points", value = TRUE)
    ),
    
    mainPanel(
      plotlyOutput("unempPlot")
    )
  )
)

server <- function(input, output) {
  
  filtered_data <- reactive({
    t8_unemp_age %>%
      filter(
        Age %in% input$age_select,
        Year >= input$year_range[1],
        Year <= input$year_range[2]
      )
  })


t8_unemp_age <- clean8_data %>%
  mutate(
    Labor_Force = Total + Unemp,
    Unemployment_Rate = Unemp / Labor_Force
  ) %>%
  group_by(Year, Age) %>%
  summarise(
    Unemployment_Rate = mean(Unemployment_Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Age, Year) 


p <- ggplot(t8_unemp_age, aes(
      x = Year,
      y = Unemployment_Rate,
      color = Age,      
      group = Age,      
      text = paste(     
        "Year:", Year,
        "<br>Age:", Age,
        "<br>Unemployment Rate:", percent(Unemployment_Rate, accuracy = 1)
      )
    )) +
  geom_line(linewidth = 1.2, alpha = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "Unemployment Rate by Age Group (2011–2024)",
    subtitle = "Comparing unemployment trends across age groups",
    x = "Year",
    y = "Unemployment Rate",
    color = "Age Group"   
  ) +
  scale_x_continuous(breaks = 2011:2024) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(values = myColors) +
  theme_fivethirtyeight() +
  theme(
    axis.title = element_text(),
    text = element_text(family = "Times New Roman")
  )

# interactive with hover feature
ggplotly(p, tooltip = "text") %>% 
  layout(
    hoverlabel = list(
      font = list(
        family = "Times New Roman",  
        size = 12,                  
        color = "black"              
      ),
      bgcolor = "white",             
      bordercolor = "transparent"
    ),
    xaxis = list(fixedrange = TRUE), 
    yaxis = list(fixedrange = TRUE)
  ) %>%
  config(displayModeBar = FALSE)

```




```{r Dareens Graph Section}

```

```{r Sarahs Graphs Section}

```

```{r Sammys Graph Section}

```

```{r Rishitas Graph Section}

```

```{r Zuwidyas graph section}

```

