---
title: 'Aging Job Prospects: 2021–2023 Analysis'
author: "Zuwiyda Hirse"
date: "2026-02-23"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)

# Make a function that cleans ONE year
clean_one_year <- function(file, year_num) {

  # read the sheet, but skip the title rows
  raw <- read_excel(file, skip = 2, col_names = FALSE)

  #row 1 and row 2 are headers inside the data
  header1 <- as.character(unlist(raw[1, ]))
  header2 <- as.character(unlist(raw[2, ]))

  # use header2 when it exists, otherwise use header1
  col_names <- ifelse(!is.na(header2) & header2 != "", header2, header1)

  # remove the two header rows and set the column names
  df <- raw[-c(1, 2), ]
  names(df) <- col_names

  #clean column names
  df <- df %>% clean_names()

  # keep only real rows with an occupation name
  df <- df %>% filter(!is.na(occupation), occupation != "")

  # turn all the age columns into numbers (fixes "-" and weird text)
  df <- df %>%
    mutate(across(-occupation, ~ parse_number(as.character(.x))))

  # add year column
  df <- df %>% mutate(year = year_num)

  return(df)
}

#Clean each year
d21 <- clean_one_year("cps_2021_11b.xlsx", 2021)
d22 <- clean_one_year("cps_2022_11b.xlsx", 2022)
d23 <- clean_one_year("cps_2023_11b.xlsx", 2023)

# Stack them together
bls_clean <- bind_rows(d21, d22, d23)
glimpse(bls_clean)

```

This R Markdown document is made interactive using Shiny. Unlike the more traditional workflow of creating static reports, you can now create documents that allow your readers to change the assumptions underlying your analysis and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



