---
title: "JobAgeProject"
author: "Obydah Jazaeri, Sarah Rosen"
date: "2026-02-21"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
<<<<<<< HEAD
library(shiny)
library(readxl)
library(tidyverse)

knitr::opts_chunk$set(echo = FALSE)

jobdata <- read_excel("cpsa2024.xlsx")
=======

library(janitor)    
library(readxl)
library(dplyr)
library(tidyr)
library(stringr) 
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
data2017 <- read_excel("cpsaat11b.xlsx")
data2018 <- read_excel("data2018.xlsx")
data2019 <- read_excel("data2019.xlsx")
>>>>>>> f9aaddcf96db12d53e5ed7ab45988ad945cc5f9a
```

```{r cleaning}
clean_cps_year <- function(df, year) {
  
  df %>%
    slice(-c(1:3)) %>%   # remove header junk
    rename(
      occupation = 1,
      median_age = 2,
      `16_to_19_years` = 3,
      `20_to_24_years` = 4,
      `25_to_34_years` = 5,
      `35_to_44_years` = 6,
      `45_to_54_years` = 7,
      `55_to_64_years` = 8,
      `65_years_and_over` = 9
    ) %>%
    filter(occupation != "Total employed") %>%
    
    # DROP median age before pivoting
    select(-median_age) %>%
    
    pivot_longer(
      cols = -occupation,
      names_to = "age_group",
      values_to = "employment_thousands"
    ) %>%
    mutate(
      year = year,
      employment_thousands = readr::parse_number(employment_thousands)
    )
}

```

<<<<<<< HEAD
```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

## 2019 11b Data

Represents employment by age group, given a drop-down menu for individual age groups.
```{r import, echo=FALSE}
df_2019 <- read_excel("cpsaat11b-2019.xlsx", skip = 4)
=======


>>>>>>> f9aaddcf96db12d53e5ed7ab45988ad945cc5f9a

head(df_2019)

colnames(df_2019)

colnames(df_2019)[1] <- "Occupation"

colnames(df_2019) <- gsub("\\r\\n", " ", colnames(df_2019))

colnames(df_2019)

df_2019 <- df_2019[, -c(2, ncol(df_2019))]

df_2019_tidy <- df_2019 %>%
  pivot_longer(
    cols = -Occupation, 
    names_to = "age_group",
    values_to = "employed"
  ) %>%
  mutate (
    year = 2019,
    employed = as.numeric(employed)
  )
glimpse(df_2019_tidy)
```

```{r, echo=FALSE}
inputPanel(
  selectInput("age", "Select Age Group:",
              choices = unique(df_2019_tidy$age_group))
)

renderPlot({
  df_2019_tidy %>%
    filter(age_group == input$age) %>%
    slice_max(employed, n = 10) %>%
    ggplot(aes(x = reorder(Occupation, employed),
               y = employed)) +
    geom_col() +
    coord_flip()
})
```


