---
title: "JobAgeProject"
author: "Obydah Jazaeri, Sarah Rosen"
date: "2026-02-21"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
library(shiny)
library(readxl)
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

jobdata <- read_excel("cpsa2024.xlsx")
```



```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

```{r}
df_2019_raw <- read_excel("cpsaat11b-2019.xlsx",
                          skip = 1,
                          col_names = FALSE)
```

```{r}
colnames(df_2019_raw) <- as.character(df_2019_raw[2, ])
df_2019_raw <- df_2019_raw[-c(1,2), ]
```


Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



